---
title: "Bayesian Learning"
subtitle: Lab 3
author: "Emil K Svensson and Rasmus Holm"
date: '`r Sys.Date()`'
output:
    pdf_document:
        includes:
            in_header: styles.sty
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1

```{r}
rainfall <- read.table("../data/rainfall.dat", header=F)
```

Prior

\begin{align*}
\mu &\sim N(\mu_{0}, \tau_{0}^{2}) \\
\sigma^2 &\sim \text{Inv--} \chi^2(\nu_{0}, \sigma_{0}^{2})
\end{align*}

Likelihood

\begin{align*}
\mathbf{y} | \mu, \sigma^2 &\sim N(\mu, \sigma^2)
\end{align*}

Posterior

\begin{align*}
\mu | \sigma^2, \mathbf{x} &\sim N(\mu_{n}, \tau_{n}^{2}) \\
\sigma^2 | \mu, \mathbf{x} &\sim \text{Inv--} \chi^2 \left( \nu_{n}, \frac{\nu_{0} \sigma_{0}^{2} + \sum_{i=1}^{n} (x_{i} - \mu)^2}{n + \nu_{0}} \right)
\end{align*}

where (?)

\begin{align*}
\mu_{n} &= \frac{\frac{1}{\tau_{0}^{2}}}{\frac{1}{\tau_{0}^{2}} + \frac{n}{\sigma^2}} \mu_{0} +
\frac{\frac{n}{\sigma^2}}{\frac{1}{\tau_{0}^{2}} + \frac{n}{\sigma^2}} \bar{y} \\
\frac{1}{\tau_{n}^{2}} &= \frac{1}{\tau_{0}^{2}} + \frac{n}{\sigma^2} \\
\nu_{n} &= \nu_{0} + \frac{n}{\sigma^2}
\end{align*}

## a)

```{r}


mun <- function(y,mu0,tau0sq){
  n <- length(y)
  pt1 <- ((1 / tau0sq) / ((1 / tau0sq) + n) ) * mu0
  pt2 <- (n / ( (1/tau0sq) + n)) * mean(y) 
  
  pt1 + pt2
}


```


## b)

## c)

\newpage

# Question 2

## a)

## b)

## c)
